# Week 2, Linear Regression

This week's exercise is about regression analysis. Tasks include:

* data import and simple manipulation 
* data exploration 
* single and multiple predictor linear regression analysis including model diagnostics and interpretation.

Source data: [http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt](http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt).

## Starting routines and getting to know data

```{r}
#Setting working path
setwd("/home/ls/R/projekteja/IODS-project/")

#Reading saved datafile as "excercise2"
exercise2_data <- read.table("./data/ex2.rData")

#Structure of data set (166 obs, 7 vars)
str(exercise2_data)

#First 10 observations (showing single string variable, 6 continuous variables), 
head(exercise2_data,n=10)

summary(exercise2_data)

#Gender proportions: females 66.3%, males 33.7%
prop.table(table(exercise2_data$gender))
```
So that was basic information about data set, which consist of 166 observations and single factor, six continuous variables, with 66% of subjects being female. For more information, please visit [http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-meta.txt](http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-meta.txt) where the original data set described.

## Visualizing pairwise distributions

```{r}
#Loading additional packages and plotting bivariate distributions 
library(ggplot2)
library(GGally) 

ggpairs(exercise2_data, aes(col=gender, alpha=0.3), 
        upper=list(continuous = wrap("cor", size=2.5)),
        lower=list(combo=wrap("facethist", bins=25))) +
  scale_fill_manual(values = c("red","blue"))
```

In graph above, red color indicates female gender, blue male. Correlation text size decreased for better fit.  

Highest absolute correlation coefficient with Points are:

Variables          | Pearson R
------------------ | -------------
Attitude * Points  |  0.437        
stra * Points      |  0.146      
surf * Points      | -0.144 

-> Let's select these three as predictors. Points is the outcome.

## Multiple linear regression model #1

```{r}
glm1 <- lm(Points ~ Attitude + stra + surf, data=exercise2_data)
summary(glm1)
```

Attitude is the only statistically significant predictor for Points (p<0.001).  

Surf is least significant predictor: let's remove it from the model.

## Multiple linear regression model #2

```{r}
glm2 <- lm(Points ~ Attitude + stra, data=exercise2_data)
summary(glm2)
```

Attitude is still the only statistically significant predictor for Points (p<0.001).  

Stra is not significant, but still p<0.1. We might leave the model as it is, but out of curiosity, let's remove Stra anyway and see what happens. 

## Simple linear regression model

```{r}
glm3 <- lm(Points ~ Attitude, data=exercise2_data)
summary(glm3)
```

Model parameter estimates interpretation:

Attitude is still highly significant predictor for Points (p<0.001).

Intercept equals 11.6. This is the value of Points where Attitude=0, based on model formula.

Slope parameter of Attitude equals to 0.352, i.e. single unit increase in Attitude reflects as 0.352 increase to Points.

Coefficents of determination: unadjusted 19.1%, adjusted 18.6%. Both are a bit lower than in previous model. But let's leave it that way.

## Diagnostic plots for model #3

```{r}
#Drawing diagnostic plots. Choosing the plots 1, 2 and 5. Full list of plot options:
#1. Residual vs fitted
#2. Normal QQ plot	
#3. Scale-location	
#4. Cooks distance
#5. Residuals vs. Leverage
#6. Cooks distances vs Leverage

#Defining plot matrix of 1x3. Pin defines 1:1 aspect ratio for each sub plot.
par(mfrow = c(1,3), pin=c(1.75,1.75))
plot(glm3, which=c(1,2,5))
```

Residual vs. fitted plot: Distibution is symmetric, there are no outliers. Looks fine.

Normal Quantile-Quantile plot: There is slight deviation from the diagonal at the ends. However, this is still not yet alarming at all.

Residual vs. Leverage plot: Highest Cook's distance value is <0.05, i.e. low. No problems here.

**Conclusion: model seems to be adequate, assumptions are met, based on visual examination. Attitude predicts value of Points.**
