# Week 4, Clustering and Classification

This week's exercise is about cluster analysis and classification. Tasks include:

* data exploration incl. correlations
* standardization
* linear discriminant analysis
* K-means clustering
* DM for next week data.

## 2. Starting routines and getting to know data

Setting working path, loading library and data set.

```{r}
#Setting working path
setwd("/home/ls/R/projekteja/IODS-project/")

# access the MASS package
library(MASS)

# load the data
data("Boston")
```

Boston data is included in the MASS package. Data is about housing values in suburbs of Boston city, with other demographic and environmental information as well.
Further information: https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html.

List of variables (from above url):  

Variable name  | Definition  
------------- | -------------  
**crim**  | per capita crime rate by town.  
**zn**  | proportion of residential land zoned for lots over 25,000 sq.ft.  
**indus**  | proportion of non-retail business acres per town.  
**chas**  | Charles River dummy variable (= 1 if tract bounds river; 0 otherwise).  
**nox**  | nitrogen oxides concentration (parts per 10 million).  
**rm**  | average number of rooms per dwelling.  
**age**  | proportion of owner-occupied units built prior to 1940.  
**dis**  | weighted mean of distances to five Boston employment centres.  
**rad**  | index of accessibility to radial highways.  
**tax**  | full-value property-tax rate per \$10,000.  
**ptratio**  | pupil-teacher ratio by town.  
**black**  | 1000(Bk - 0.63)$^{2}$ where Bk is the proportion of blacks by town.  
**lstat**  | lower status of the population (percent).  
**medv**  | median value of owner-occupied homes in \$1000s.  

Sources:  
- _Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. J. Environ. Economics and Management 5, 81–102._  
- _Belsley D.A., Kuh, E. and Welsch, R.E. (1980) Regression Diagnostics. Identifying Influential Data and Sources of Collinearity. New York: Wiley._



Loading library dplyr for further DM needs.
```{r}
library(dplyr) 

#Data dimensions, variable names and some values.
glimpse(Boston)
summary(Boston)
```
The Boston data frame has 506 rows and 14 columns. All data is numeric. Variables are continuous expect binary variable **chas**, _("1 if tract bounds Charles River; 0 otherwise")_. Response/target variable will be **crim** _("per capita crime rate by town")_, which has range 0.06-89.0 and mean=3.6 & median=0.26. Difference between mean and median suggests strongly skewed distribution.

## 3. Graphical overview and summaries

Descriptives were already shown in step 2. Variable distributions have clearly distinct variances and locations. Some variables seem to be strongly skewed as well. Let's have a visual.

```{r}
# Bivariate scatter plots (excluding 4th variable chas)

par(mfrow = c(1,2), pin=c(1.75,1.75))
hist(Boston$crim,col="blue",xlab=NULL)
boxplot(Boston$crim,col="blue", main="Boxplot of Boston$crim")
```

We can see that crime distribution is strongly skewed. We are going to standardize and categorize variables later. 

Pairwise distributions (the plot is enlarged with defining additional R-chunk parameters, which are not shown in output):

```{r fig1, out.width = '100%', fig.asp=1}
# Bivariate scatter plots (excluding 4th variable chas)
pairs(Boston[,c(1:3,5:dim(Boston)[2])], pch=19, cex=0.05, lower.panel=NULL)
```

Scatter plot matrix shows that bivariate distributions are not always multinormal (or almost never, to be honest). Some distributions do have large empty areas and/or outlier observations, like **rad** _(index of accessibility to radial highways)_. Not good.

Some mutual bivariate associations seems to be stronger than others. Let's calculate correlations.

```{r}
# calculate the correlation matrix and round it
cor_matrix<-cor(Boston) %>% round(digits=2)

# print the correlation matrix
cor_matrix
```

These are Pearson correlation coefficients, which are not always reliable in a case of non-normality. But let's keep them anyway, like in Datacamp. However, it's challenging to assimilate so many numbers. Better to have a visualization.

```{r}
# installing and/or loading corrplot library for correlation coefficient visualization
if (!require("corrplot")) {
   install.packages("corrplot")
   library(corrplot)
}

# visualize the correlation matrix
corrplot.mixed(cor_matrix, tl.cex=0.75, number.cex=0.75, number.digits=2)
```

That's fancy way to present correlation coefficient. We can directly see that there are some strong correlations indicated by big spheres, blue ones for negative and red ones for positive values. Highest value is +0.91, which is between **rad** and **tax** (_index of accessibility to radial highways_ and _full-value property-tax rate per \$10,000_).

## 4. Standardization and categorization

```{r fig2, out.width='100%', fig.asp = 1}
# center and standardize variables
# 'as.data.frame' is needed since it makes referring variables easier later
boston_scaled <- as.data.frame(scale(Boston))

# Bivariate scatter plots
pairs(boston_scaled, pch=19, cex=0.05, lower.panel=NULL)

#Means, second argument refers to columns
apply(boston_scaled, 2, FUN=mean)

#Variances, second argument refers to columns
apply(boston_scaled, 2, FUN=var)
```

We can see that now all variables has been standardized, i.e. scaled to have mean=0 (or very close to it, at least) and variance=1. This can't heal difficult distributions like **rad** or **chas** has, though. Correlations are the same, so they are not shown here again.

Note: For some reason scaling must be for R data.frame object, otherwise further operations won't work. This means that command  
&nbsp;&nbsp;&nbsp; _boston_scaled <- scale(Boston)_  
is not enought, but I need to put it this way:  
&nbsp;&nbsp;&nbsp; _boston_scaled <- as.data.frame(scale(Boston))_.  
So strange... And the same applies for task #7.

### Categorization

Creating categorized version of crime variable. Cut points are quantiles, so number of categories will be four.

```{r}
# create a quantile vector of crim and print it
bins <- quantile(boston_scaled$crim)
bins

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks=bins, include.lowest=TRUE, labels=c("low","med_low","med_high","high"))

# look at the table of the new factor crime
table(crime)
```

Looks fine, distribution is as close to 25/25/25/25 percentages as possible.

```{r}
# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)
```

Now original **crim** has been replaced to categorical **crime**.

### Dividing the dataset to train and test sets

```{r}
# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

dim(train)
dim(test)
```

Now datasets **train** and **test** has been created, with the first comprising 80% and latter 20% of original cases. 

## 5. Linear discriminant analysis

Fitting LDA model on the train set with crime rate categorizations as target variable. Other variables are used as predictors.

```{r}
# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)

# print the lda.fit object
lda.fit

classes <- as.numeric(train$crime)

# LDA (bi)plot
plot(lda.fit, col=classes)
```

Interestingly plot() function adds group names as data value symbols independently on pch parameters. For example, following commands are yielding identical plots:  
- *plot(lda.fit, col=classes, pch=19)*  
- *plot(lda.fit, col=classes, pch=classes)*  
I have no clue why. Showing category value as symbols looks very awkward.

Anyway, plot visualizes how target variable classes are separated by the linear combinations of predictor variables. Looks like crime quantile groups are mostly very nicely separated. _High_ crime group seems to be most consistent while _med_high_ is most scattered around. 

Let's enhance plot by adding arrows into it, just like in Datacamp exercise. 

```{r}
# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

plot(lda.fit, col=classes, pch=classes, dimen=2)
lda.arrows(lda.fit, myscale = 2)
```

Looks like **rad** _("index of accessibility to radial highways")_ is an important factor here. Watch out for living close to radial highways!

## 6. Predicting crime with LDA model

```{r}
# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)
```
Cross-tabulation shows that in 13+17+15+20 cases (which is `r 13+17+15+20`) prediction is fully correct, leaving out `r dim(test)[1] -13-17-15-20` more or less incorrect cases. Proportion of correct predictions is hence `r (13+17+15+20)/dim(test)[1]`. We can see that if prediction has gone wrong, it has still has been mostly placed into category close to the real one. And, we can see that all 20 high crimes cases are correctly predicted. Not bad at all, I guess.

## 7. Distances, K-means clustering

### Euclidian distances

```{r}
# reloading the Boston dataset 
data('Boston')

# standardizing the dataset
# again, as.data.frame is needed, otherwise further operations won't work.
boston_scaled <- as.data.frame(scale(Boston))

# euclidean and manhattan distance matrix
dist_eu <- dist(boston_scaled)
dist_man <- dist(boston_scaled, method="manhattan")

# look at the summaries of the distances
summary(dist_man)
summary(dist_man)

par(mfrow = c(2,2), pin=c(1.75,1.75))
#Euclidean
hist(dist_eu,col="blue",xlab=NULL)
boxplot(dist_eu,col="blue", main="Boxplot of dist_eu")
#Manhattan
hist(dist_man,col="blue",xlab=NULL)
boxplot(dist_man,col="blue", main="Boxplot of dist_man")
```

Mean value for Euclidean distance is around 5, range is 0.1 - 14.4. Distribution is slightly skewed to the right.

Mean value for Manhattan distance is around 14, range is 0.3 - 48.9. Distribution is a bit more skewed to the right.

### K-means clustering

Let's conduct k-means clustering with three clusters, which, I think, might be good low-but-not-super-low number for centers/clusters.

```{r fig3, out.width = '100%', fig.asp=1}
# k-means clustering with three clusters (=semirandomly selected number)
km <-kmeans(boston_scaled, centers=3)

# plot the Boston dataset with clusters
pairs(boston_scaled, col=km$cluster, pch=19, cex=0.05, lower.panel=NULL)
```

Three cluster solutions might be plausible. Groups are formed nicely. But let's see which is optimal number based on calculations.


```{r}
# Setting a seed for random generator
set.seed(322654435)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_scaled, k)$tot.withinss})

# visualize the results
library(ggplot2)
ggplot() + 
  geom_line(aes(x=1:k_max, y=twcss)) +
  scale_x_continuous(breaks=c(1:10)) +
  xlab("Number of cluster")
```

If the principle is that _the optimal number of clusters is when the value of total WCSS changes radically_, then two clusters would be good choice now. After two clusters _total within sum of squares_ decreased slower. So let's conduct k-means clustering again, now with two centers, and visulize results.

```{r fig4, out.width = '100%', fig.asp=1}
# k-means clustering
km2 <-kmeans(boston_scaled, centers=2)

# plot the Boston dataset with clusters
pairs(boston_scaled, col=km2$cluster, pch=19, cex=0.05, lower.panel=NULL)
```

This looks plausible as well. Dot colors indicating cluster seem to mostly create distinct groups in these subplots.

## Bonus task

Reloading and scaling Boston data set. AFAIK, this is not needed to done again, but let's follow instructions. K-means clustering with three center. Fitting LDA with cluster as a target variable, keeping all Boston variables as predictors. Adding arrows with previously created custom function, replacing default red color with blue for better definition. Aspect ratio, text size and arrow length scaling changed from default values as well.

```{r fig5, out.width = '75%', fig.asp=1}

# reloading the Boston dataset 
data('Boston')
boston_scaled <- as.data.frame(scale(Boston))

# k-means clustering with three centers
km3 <- kmeans(boston_scaled, centers=3)
km3

# LDA with cluster as a target
lda2.fit <- lda(km3$cluster ~ ., data=boston_scaled)

# print the lda.fit object
lda2.fit

# target classes as numeric
classes <- as.numeric(km3$cluster)

# plot the lda results
plot(lda2.fit, col=classes, pch=classes)

# Using previously defined function for lda biplot arrows
lda.arrows(lda2.fit, color="blue", tex=0.8, myscale=5)
```

Looks like variables **nox** _(nitrogen oxides concentration, parts per 10 million)_, **tax** _(full-value property-tax rate per \$10,000)_ and **zn** _(proportion of residential land zoned for lots over 25,000 sq.ft)_ are the most influental separators for the clusters among all variables, based on arrow lengths. 

## Super-Bonus task

Running the given code for the scaled train data. The code creates a matrix product, which is a projection of the data points. Installing and loading plotly package and creating 3D plots.

Adjusting the code. Defining symbol color as train data set crime class. Drawing another 3D plot where the color is defined by the clusters of the k-means. Plots are only shown in RStudio viewer, so no output in course diary.

```{r}
# creating train set (again)
train <- boston_scaled[ind,]

#K-means clustering with three centers:
km4 <- kmeans(train, centers=3)

#restoring original crime classification into train data
crime <- cut(train$crim, breaks=bins, include.lowest=TRUE, labels=c("low","med_low","med_high","high"))

#model_predictors <- dplyr::select(train, everything())
model_predictors <- dplyr::select(train, c(-crim))
#dim(model_predictors)

# check the dimensions
dim(model_predictors)
dim(lda.fit$scaling)

# matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)

# installing and/or loading plotly package for 3D plot
if (!require("plotly")) {
   install.packages("plotly")
   library(plotly)
}

#Create a 3D plot (Cool!) of the columns of the matrix product by typing the code below.
#plot_ly(x=matrix_product$LD1, y=matrix_product$LD2, z=matrix_product$LD3, type='scatter3d', mode='markers')

#Graph 1
plot_ly(x=matrix_product$LD1, y=matrix_product$LD2, z=matrix_product$LD3, color=crime, type='scatter3d', mode='markers') 

#Graph 2
plot_ly(x=matrix_product$LD1, y=matrix_product$LD2, z=matrix_product$LD3, color=as.factor(km4$cluster), type='scatter3d', mode='markers')

```

How do plots differ? Well, data points are identical in x-y-z space. Only number of categories/colors (4 vs. 3) differ, so do group distribution. After all, both plots are telling pretty much the same story: there is one clearly distinct blob (upper quartile of crime rates) while rest of dots are somewhat grouped as well. Maybe there is bit more overlapping (variation) in crime category plot.
